---
title: "Community Phylogenetics Tutorial"
author: "Farah Sadoon"
date: "2025-11-24"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Loading libraries
library(ape)
library(picante)
library(pez)
library(phytools)
library(vegan)
library(adephylo)
library(phylobase)
library(geiger)
library(mvMORPH)
library(OUwie)
library(hisse)
library(BAMMtools)
library(phylosignal)
library(Biostrings)
library(devtools)
library(ggplot2)
library(kableExtra)
library(betapart)
library(gridExtra)
library(reshape2)
library(ggtree)

# Download files from Marc Cadotte's and Jonathan Davies' book 
# "Phylogenies in Ecology: A guide to concepts and methods":

#dir.create("../data/Jasper")

#download.file("https://raw.githubusercontent.com/pedrohbraga/PhyloCompMethods-in-R-workshop/master/data/Jasper/resources/data/Jasper/jasper_data.csv", 
#              "../data/Jasper/jasper_data.csv")

#download.file("https://raw.githubusercontent.com/pedrohbraga/PhyloCompMethods-in-R-workshop/master/data/Jasper/resources/data/Jasper/jasper_tree.phy", 
#              "../data/Jasper/jasper_tree.phy")

# Download the tree files from this paper

# download.file("https://onlinelibrary.wiley.com/action/
# downloadSupplement?doi=10.1111%2Fj.1461-0248.2009.01307.x&attachmentId=179085359", 
#              destfile = "data/ele_1307_sm_sa1.tre")

# If the download of the above file does not work, 
# download the first supplementary material from the following paper
# and place it within your 'data' directory:
# https://onlinelibrary.wiley.com/doi/abs/10.1111/j.1461-0248.2009.01307.x
```

## A brief introduction to phylogenies
Phylogenetic trees (i.e., evolutionary tree or cladogram) are branching diagrams illustrating the evolutionary relationships among taxa.

Phylogenetic trees can be constructed based on species’ genes and/or on traits. Phylogenetic reconstruction through analyses of molecular sequences usually consists of three distinct procedures: (i) sequence alignment; (ii) character coding; and (iii) tree building. There is a range of available options for each of these steps, with no simple objective method for choosing between them.

The root of a phylogenetic tree indicates that an ancestral lineage gave rise to all organisms on the tree. A branch point indicates where two lineages diverged. A lineage that evolved early and remains unbranched is a basal taxon. When two lineages stem from the same branch point, they are sister taxa. A branch with more than two lineages is a polytomy. Extracted from: https://doi.org/10.1016/j.tree.2012.10.015 
In this course, we will not cover the steps necessary to reconstruct phylogenies. We will assume that a phylogenetic relationship is already available for the taxa of interest.


## `ape` and the `phylo` object in R
The core of phylogenetics analyses in R is the library ape - an acronym that stands for Analysis of Phylogenetics and Evolution.

From `ape`, we depict two important functions to read and load trees into R: `read.nexus()` and `read.tree()`

#### `read.nexus()` reads NEXUS formatted trees, which is organized into major units known as blocks and usually have the following basic structure:

<pre>
#NEXUS

[!This is the data file used for my dissertation]

BEGIN TAXA;
    TaxLabels Scarabaeus Drosophila Aranaeus;

END;

BEGIN TREES;
    Translate beetle Scarabaeus, fly Drosophila, spider Aranaeus;
    Tree tree1 = ((1,2),3);
    Tree tree2 = ((beetle,fly),spider);
    Tree tree3 = ((Scarabaeus,Drosophila),Aranaeus);
END;
</pre>


#### `read.tree` reads Newick formatted trees, which commonly have the following format:
<pre>
(Bovine:0.69395,(Hylobates:0.36079,(Pongo:0.33636,(G._Gorilla:0.17147, (P._paniscus:0.19268,H._sapiens:0.11927):0.08386):0.06124):0.15057):0.54939, Rodent:1.21460)
</pre>

When phylogenies are created, loaded and handled in `ape`, their objects are represented as a list of *class* `phylo`.

Let’s try reading the following text string of a tree (is it in Newick or in NEXUS format?) and see what R tells us when we call it:

```{r}
tree.primates <- read.tree(text="((((Homo:0.21, Pongo:0.21):0.28, Macaca:0.49):0.13, 
                           Ateles:0.62):0.38, Galago:1.00);")
tree.primates
```

A class phylo object distributes the information of phylogenetic trees into six main components:

```{r}
str(tree.primates)
```
Since `phylo` is a list, we can access its elments by using the `$` symbol. We can access the matrix `edge` by running `phylo_object$edge`, which has the beginning and ending node number for all the nodes and tips of a given tree. In summary, this allows us to keep track of the internal and external nodes of the tree: 

```{r}
tree.primates$edge
```

